Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-09-03 manu-draft.Rmd
draft discussion done, need to add refs

The observed conditions in Tampa Bay from April to July, 2021 following effluent release from Piney Point together represent an adverse environmental response to a large pulse of inorganic nitrogen into the system.  Collectively, these observations provide a weight of evidence that conditions in 2021 were anomalous when compared to the long-term record of water quality and habitat data for Tampa Bay.  These anomalous events included 1) a large, spring diatom bloom in the vicinity of the discharge at Port Manatee, 2) high abundance of the macroalgae *Lyngbya sp.* in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay, and 4) high incidence of fish kill reports prompting local governments to remove over 1800 tons of dead fish from shoreline areas.  The observed water quality conditions throughout this period, particularly for total nitrogen, chlorophyll-a, and secchi depth, were outside of normal ranges for an unusual amount of observations supported by the decades of long-term monitoring data as a baseline for the region (Figures \@ref(fig:wqmap) and \@ref(fig:wqtrnds), Table \@ref(tab:wqsumtab)).  Although these out-of-range results had specific temporal and spatial patterns during the period of observation, they were seen throughout the bay suggesting a system-wide effect of effluent from Piney Point.  In addition to the changing environmental conditions observed during the summer, the Piney Point event is also anomalous in the volume of effluent entering the system and estimated nutrient load to lower Tampa Bay.  Spill events reported to FDEP have been documented for years (e.g., industrial spills, stormwater pump failures, sanitary sewer overflows) and spill volumes for these for these events are small (median volume 3630 gallons, [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 285 million gallons discharged from Piney Point. Moreover, the estimated nutrient load of 205 tons of nitrogen to Tampa Bay from Piney Point discharged over the ten day period exceeded current annual estimates of the external load into lower Tampa Bay.  From both a monitoring perspective and the nature of the effluent into Tampa Bay, the observed response of conditions in the Bay in 2021 is an outlier compared to past years. These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator and asset for environmental quality.

@@

Potential short and long-term impacts to Tampa Bay resources beyond water quality, macroalgae, and seagrasses may also have occurred.  Fisheries resources have undoubtedly been impacted by *K. brevis* blooms in June and July as documented by the fish kill reports (Figure \@ref(fig:redtide)).  Quantitative data on changes in abundance and diversity of fish species in Tampa Bay following the red tide event are forthcoming. Routine sampling by the Fisheries Independent Monitoring Program of the FWC occurs annually in Tampa Bay and a long-term record back to 1998 provides detailed information on species richness and abundance for the major segments of the bay. @Schrandt21 developed a Tampa Bay Nekton Index to assist with fisheries management strategies for the bay and demonstrated a sensitivity of the index to a significant red tide event in lower Tampa Bay in 2005. Given the observed red tide concentration in 2021 and the magnitude of reported fish kills, a similar response in the fish community as reflected by the TBNI is expected.  Of note is that fisheries resources improved quickly following surveys after 2005, although temporary recreational or commercial closures may need to be implemented to allow the fish community to recover.  As of writing, catch and release only limits have been extended through the fall for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*).  @SWitzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following dishcarge to the subembayment and compared the results to fisheries data for years prior.  Fish community structure and species composition did not differ compared to the pre-impact period. From 2003 to 2007, an estimated 1.9 billion gallons was discharged to Bishop Harbor, where various treatment methods were used to mitigate potential impacts of introducing acidic and nutrient-rich process water to surface waters. @Garrett11 noted bloom concentrations of HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K. brevis* and dense diatom blooms, in Bishop Harbor during this time.

Additional monitoring was conducted for benthic communities to sample sediment quality at Port Manatee and surrounding waters in April 2021.  Biological surveys were used to calculate scores for the Tampa Bay Benthic Index (TBBI) as a measure of the ecological health of the benthic community based on the quality of the sediments they inhabit [@tbep0106]. Similar to other reporting tools for Tampa Bay, the TBBI categorizes sediment samples into "good", "fair", and "poor" as a qualitative summary of enumerated taxonomic data that can support management decisions. Results from initial sampling suggest a mix of conditions with dominated by "fair" and "good" samples near Port Manatee reflecting the high spatial variability of macroinvertebrate communities in estuarine sediments.  Comparison of the initial 2021 samples to historical benthic conditions using TBBI scores for comparison suggested relatively consistent sediment quality back to earlist sampline in 1993.  Additional sampling is planned at the same sites surveyed in April to further assess potential impacts of Piney Point on sediment quality. Finally, effects of changing environmental conditions and red tide on marine mammals (e.g., cetaceans, sirenians) was also a concern given their abundances in the bay and use of bay resources.  Detailed baseline information and data relative to 2021 are lacking for these species relative to the primary data reported herein.  At least one [report](https://myfwc.com/research/manatee/rescue-mortality-response/statistics/mortality/2021/) to FWC of a manatee mortality in Tampa Bay at the peak of the red tide bloom in July was recorded, although cause of death was undetermined. Regardless, it is reasonable to assume toxicity of *K. brevis* can have a negative effect on regional populations of large mammals.  This is of particular concern given the recent report of manatee mass die-offs on the east coast of Florida related to seagrass losses.

* Comparison to other locations/past events - Grand Bay, Bishop Harbor, Huelva estuary [@Perez10;@Perez16], Dillon report about Grand Bay [link](https://www.wrri.msstate.edu/pdf/2016dillon_finalreport.pdf)
* Analysis limitations: no smoking gun but 2021 is an anomaly, additional info (benthic diversity TBD, nekton diversity TBD, large mammals, etc.), response-based monitoring may be biased
* Long-term closure plan?  NGS-N is treated with spray evaporation system, but concentrations of TN/TP higher than NGS-S (see EO)/
* Current challenges in TB/southwest FL - OTB, seagrass loss (possibly linked to 2018 red tide, effects of Hurricanes [@Tomasko20]), red tide, climate change
* Risk of decline (IRL ex.), regression of past progress

>>>>

The observed conditions in Tampa Bay from April to July, 2021 following effluent release from Piney Point together represent an adverse environmental response to a large pulse of inorganic nitrogen into the system.  Collectively, these observations provide a weight of evidence that conditions in 2021 were anomalous when compared to the long-term record of water quality and habitat data for Tampa Bay.  These anomalous events included 1) a large, spring diatom bloom in the vicinity of the discharge at Port Manatee, 2) high abundance of the macroalgae *Lyngbya sp.* in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay, and 4) high incidence of fish kill reports prompting local governments to remove over 1800 tons of dead fish from shoreline areas.  The observed water quality conditions throughout this period, particularly for total nitrogen, chlorophyll-a, and secchi depth, were outside of normal ranges for an unusual amount of observations supported by the decades of long-term monitoring data as a baseline for the region (Figures \@ref(fig:wqmap) and \@ref(fig:wqtrnds), Table \@ref(tab:wqsumtab)).  Although these out-of-range results had specific temporal and spatial patterns during the period of observation, they were seen throughout the bay suggesting a system-wide effect of effluent from Piney Point.

In addition to the changing environmental conditions observed during the summer, the Piney Point event is also anomalous in the volume of effluent entering the system and estimated nutrient load to lower Tampa Bay.  Spill events reported to FDEP (e.g., industrial spills, service line failures, sanitary sewer overflows) provide additional context for Piney Point relative to other potential anomalous discharges to Tampa Bay. An assessment of over 800 reports to FDEP for the Tampa Bay watershed over the last five years showed spill volumes for these events are small (median volume 3630 gallons, [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 285 million gallons discharged from Piney Point. Moreover, the estimated nutrient load of 205 tons of nitrogen to Tampa Bay from Piney Point discharged over the ten day period exceeded current annual estimates of the external load into lower Tampa Bay.  From both a monitoring perspective and the nature of the effluent into Tampa Bay, the observed response of conditions in 2021 is an outlier compared to past years. These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator of environmental quality and an asset for biological resources in Tampa Bay.

@@

Potential short and long-term impacts to Tampa Bay resources beyond water quality, macroalgae, and seagrasses may also have occurred.  Fisheries resources have undoubtedly been impacted by *K. brevis* blooms in June and July as documented by the fish kill reports to FWC (Figure \@ref(fig:redtide)).  Quantitative data on changes in abundance and diversity of fish species in Tampa Bay following the red tide event are forthcoming. Routine sampling by the FWC Fisheries Independent Monitoring Program occurs annually in Tampa Bay and a long-term record back to 1998 provides detailed information on species richness and abundance for the major segments of the bay. @Schrandt21 developed a Tampa Bay Nekton Index to assist with fisheries management strategies for the bay that demonstrated a sensitivity of fisheres resrouces to a significant red tide event in lower Tampa Bay in 2005. Given the observed red tide concentration in 2021 and the magnitude of reported fish kills, a similar response in the fish community as reflected by the TBNI is expected.  Of note is that fisheries resources improved quickly following surveys after 2005, although temporary recreational or commercial closures may need to be implemented to allow the fish community to recover.  As of writing, catch and release only limits for 2021 have been extended through the fall for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*).  For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following dishcarge to the subembayment and compared the results to fisheries data for years prior.  Fish community structure and species composition did not differ compared to the pre-impact period. From 2003 to 2007, an estimated 1.9 billion gallons was discharged to Bishop Harbor, where various treatment methods were used to mitigate potential impacts of introducing acidic and nutrient-rich process water to surface waters. @Garrett11 noted bloom concentrations of HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K. brevis* and dense diatom blooms, in Bishop Harbor during this time.

Additional monitoring and evaluation of results may also be required to fully understand the long-term impacts to bay resources beyond water quality.  Sampling was conducted for benthic communities to evaluate sediment quality at Port Manatee and surrounding waters in April 2021.  Biological surveys were used to calculate scores for the Tampa Bay Benthic Index (TBBI) as a measure of the ecological health of the benthic community based on the quality of the sediments they inhabit [@tbep0106]. Similar to other reporting tools for Tampa Bay, the TBBI categorizes sediment samples into "good", "fair", and "poor" as a qualitative summary of enumerated taxonomic data that can support management decisions. Results from initial sampling suggested a mix of conditions with dominated by "fair" and "good" samples near Port Manatee, reflecting the high spatial variability of macroinvertebrate communities in estuarine sediments.  Comparison of the initial 2021 samples to historical benthic conditions using TBBI scores for comparison suggested relatively consistent sediment quality back to earliest sampling in 1993.  Additional sampling is planned later in 2021 at the same sites surveyed in April to further assess potential impacts of Piney Point on sediment quality. Finally, effects of changing environmental conditions and red tide on marine mammals (e.g., cetaceans, sirenians) was also a concern given their abundances in the bay and use of bay resources within the broader spatial distribution of these species.  Twenty preliminary manatee (*Trichechus manatus latirostris*) mortalities were [reported](https://myfwc.com/media/25649/2021preliminaryredtide.pdf) within the known red tide boundary of the impacted counties of Tampa Bay through August 2021. This is of particular concern given the recent [unusual mortality events](https://myfwc.com/research/manatee/rescue-mortality-response/ume/) for Florida manatees that is suspected to be strongly linked to seagrass losses on the east coast (e.g., Indian River Lagoon) and current trends of seagrass loss for southwest Florida.

Compared to past events, the 2021 event differed from previous releases to the bay in that effluent was discharged directly at the port instead of through channels to back-bay habitats (i.e., Bishop Harbor).  The comparison to past years provided a stark contrast to the effect of wastewater discharge location that is specific to Tampa Bay, but can inform understanding of pollutant dispersal and its impact on different habitats that vary spatially and throughout the water column.  The Tampa Bay Coastal Ocean Model [@Chen19] was used to create a three-dimensional picture of simulated daily tracer particle movement and expected dilution of nutrient concentrations following release at the point of discharge.  Although these efforts initially informed monitoring to track potential plume impacts, the results suggested that the spread of the effluent was widespread throughout the bay and persisted for many months after the discharge ceased at Port Manatee.  Plume dispersal also suggested that both open-water and back-bay habitats were exposed to nutrient rich wastewater from Piney Point.  Further, estimates of tracer concentrations suggested that nutrient dilution occurred rapidly on the order of 10- to 100-fold in a matter of days.  However, diluted concentrations were considered biologically relevant given the initial concentrations of inorganic nitrogen measured in stack effluent (Table \@ref(tab:stktab)) and the background concentrations for the bay from baseline monitoring (Table \@ref(tab:wqsumtab)).

In the broader context of mining impacts to surface waters, the results herein reinforce the understanding that legacy pollutants from phosphate mining represent a serious threat to human health and the environment.  This study adds to the growing body of literature on the impacts of effluent from gypstack release to coastal waters that can inform both event-based monitoring and anticipated environmental response.  In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have also been affected by gypstack release.  Grand Bay is a 7500 hectare protected area in southern Mississippi that has historically been exposed to phosphorus-rich and highly acidic wastewater release from a defunct gypstack [@Dillon15;@Beck18].  The first spill event in 2005 occurred following failure of the retaining walls for the gypstack after a heavy rain event and the second event in 2012 occurred after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.  Following both of these events, massive fish kills were observed likely related to acute toxicity with low pH exposure with discharge of wastewater.  Unlike Piney Point, inorganic nitrogen concentrations of the effluent were low and concern of the long-term impacts focused primarily on the effects of heavy loads of orthophosphate to the system [@Dillon15].  Regardless, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].  A general trend is that legacy wastewater from past production of fertilizer is poorly maintained and long-term plans ae insufficient to dispose of pollutants that can significantly affect coastal resources.  Piney Point and Grand Bay are not isolated examples and sustained attention and regulatory oversight on efforts to close these facilities in an environmentally responsible manner are needed at the national and international level.

Limitations of our analysis are also important to note to inform future event-based monitoring and additional research to better understand acute, short-term threats to the sustained recovery of Tampa Bay.  All of the analyses are correlative based on associations between the measured water quality observations, macroalgae, and seagrass results and, as a result, may not represent explicit cause and effect mechanisms.  However, the interpretations are supported by well-established linkages between drivers of primary production in coastal waters.  The outstanding weight of evidence that 2021 is an anomaly relative to past years provides further justification that our interpretation of results is within the limits of what would be expected by more mechanistic descriptions of the effects of large nutrient pulses to Tampa Bay.  Additional data to support these results could include explicit load-based estimates for all sources entering the bay in 2021 and these estimates are forthcoming as additional context to the events of this year.  Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the wastewater, could provide a more definite description of the recycling and fate of nitrogen in the system.  There may likely never be an irrefutable link between the observations in 2021 and the wastewater from Piney Point, but this is not necessary to support future actions that ensure the continued threats of these facilities to the health of coastal resources are eliminated through collaborative efforts by local, regional, and state partners.  These efforts are critical to managing the health of Tampa Bay given past successes and the need to address ongoing threats of climate change, population growth, habitat loss, episodic weather events, and recurring pollutant sources to the bay.
